The importance of reproducibility in COVID-19 research: the case of population density and the spread of the pandemic

### 2021-05-20 R0-Density-Reanalysis/R0-Density-Reanalysis.Rmd
First draft of background, started working on tobit models; recreated data package to include a common key for the state boundaries

```{r load-packages, include=FALSE, cache=FALSE}
library(censReg) # Package to estimate tobit model
library(kableExtra) # Package for creating tables
library(Matrix) # Used to convert list of spatial weights to sparse matrix; needed for spatialprobit
library(nlme) # Used to estimate 
library(r0density) # Data package
library(sf) # Package to work with spatial information in simple features format
library(spatialprobit) # Used to estimate spatially autoregressive tobit
library(spdep) # Utilities to create spatial weights and work with spatial data
library(tidycensus) # Package to retrieve census information 
library(tidyverse) # Data carpentry and analysis tools
library(units) # Package to work with units of measurement
```

```{r invoke-data, include=FALSE}
data(county_geo)
```

@@

```{r list-of-geographical-neighbors, include=FALSE}
# Create a list of neighbors based on the polygons; the default contiguity criterion is Queen:
county_nb <- poly2nb(county_geo_clean)
```

@@

@@

## Replicating SWN

@@

```{r R0-map, echo=FALSE, fig.cap="\\label{fig:R0-map}Basic reproductive rate in US counties (Alaska, Hawaii, Puerto Rico, and territories not shown)."}
# Choropleth map of basic reproductive number
ggplot() + # Create a ggplot
  # Choropleth map of basic reproductive number
  geom_sf(data = county_geo_clean %>% # Add the simple features object to the plot
            filter(!STATE %in% c("02", "15", "72")), # Exclude Alaska, Hawaii, and Puerto Rico
          aes(fill = R),
          color = NA) +
  # Use viridis color palette: viriis color palettes are designed to print well in monochrome, are perceptually uniform, and effective for colorblindness
  scale_fill_viridis_c(name = "R0",
                       direction = -1) +
  # Overlay counties with missing values of the reproductive number in white
  geom_sf(data = county_geo_clean %>% # Add the simple features object to the plot
            filter(!STATE %in% c("02", "15", "72"), # Exclude Alaska, Hawaii, and Puerto Rico
                   R == 0), 
          fill = "white",
          color = NA) +
  # Overlay county boundaries
  geom_sf(data = county_geo_clean %>% # Add the simple features object to the plot
            filter(!STATE %in% c("02", "15", "72")), # Exclude Alaska, Hawaii, and Puerto Rico
          fill = NA,
          color = "lightgray") +
  # Add caption
  labs(caption = "Note: counties in white represent missing values of the basic reproductive number") +
  # Set theme of plot
  theme_minimal()
```

@@

Table \ref{tab:swn-results} replicates Table 1 of SWN (p. 6).

@@

@@

# Some relevant questions

@@

```{r}
table1model1 <- censReg(R ~ density_log, 
                        left = 0,
                    data = county_geo_clean)
summary(table1model1)
```

Linear mixed models of R0 and density + % private transportation
```{r}
table1model2 <- censReg(R ~ density_log + private, 
                    data = county_geo_clean)
summary(table1model2)
```

Linear mixed models of R0 and density + % private transportation + median income
```{r}
table1model3 <- censReg(R ~ density_log + private + hincome, 
                    data = county_geo_clean)
summary(table1model3)
```

Tobit models of R0 and density + % private transportation + median income + interaction of density and % private transportation
```{r}
table1model4 <- censReg(R ~ density_log*private + hincome, 
                    data = county_geo_clean)
summary(table1model4)

```

@@

Fit spatially autoregressive tobit:
```{r}
# Fit SAR Tobit
fit_sartobit <- sartobit(R ~ density_log + private + hincome_log,
                         B,
                         ndraw = 1000,
                         burn.in = 200, 
                         showProgress = TRUE,
                         data = county_geo_clean,
                         computeMarginalEffects = TRUE)
summary(fit_sartobit)
impacts(fit_sartobit)
```

>>>>

```{r load-packages, include=FALSE, cache=FALSE}
library(censReg) # Package to estimate tobit model
library(kableExtra) # Package for creating tables
library(Matrix) # Used to convert list of spatial weights to sparse matrix; needed for spatialprobit
library(nlme) # Used to estimate 
library(r0density) # Data package
library(sf) # Package to work with spatial information in simple features format
library(spatialprobit) # Used to estimate spatially autoregressive tobit
library(spdep) # Utilities to create spatial weights and work with spatial data
library(tidycensus) # Package to retrieve census information 
library(tidyverse) # Data carpentry and analysis tools
library(units) # Package to work with units of measurement
library(viridis) # Color palettes
```

```{r invoke-data, include=FALSE}
data(county_geo)
data(state_geo)
```

@@

```{r list-of-geographical-neighbors, include=FALSE}
# Create a list of neighbors based on the polygons; the default contiguity criterion is Queen:
# Counties
county_nb <- county_geo_clean %>%
  poly2nb()
# States
state_nb <- state_geo %>% 
  filter(!STATE %in% c("02", "15", "72")) %>%
  poly2nb()
```

@@

state_listw <- nb2listw(state_nb, 
                 style="W", 
                 zero.policy = TRUE)

@@

# Reproducing SWN

@@

```{r R0-map, echo=FALSE, out.width= "1\\linewidth", fig.cap="\\label{fig:R0-map}Basic reproductive rate in US counties (Alaska, Hawaii, Puerto Rico, and territories not shown)."}
# Choropleth map of basic reproductive number
ggplot() + # Create a ggplot
  # Choropleth map of basic reproductive number
  geom_sf(data = county_geo_clean %>% # Add the simple features object to the plot
            filter(!STATE %in% c("02", "15", "72")), # Exclude Alaska, Hawaii, and Puerto Rico
          aes(fill = R),
          color = NA) +
  # Use viridis color palette: viriis color palettes are designed to print well in monochrome, are perceptually uniform, and effective for colorblindness
  scale_fill_viridis_c(name = "R0",
                       direction = -1) +
  # Overlay counties with missing values of the reproductive number in white
  geom_sf(data = county_geo_clean %>% # Add the simple features object to the plot
            filter(!STATE %in% c("02", "15", "72"), # Exclude Alaska, Hawaii, and Puerto Rico
                   R == 0), 
          fill = "white",
          color = NA) +
  # Overlay county boundaries
  geom_sf(data = county_geo_clean %>% # Add the simple features object to the plot
            filter(!STATE %in% c("02", "15", "72")), # Exclude Alaska, Hawaii, and Puerto Rico
          fill = NA,
          color = "lightgray") +
  # Add caption
  labs(caption = "Note: counties in white represent missing values of the basic reproductive number") +
  # Set theme of plot
  theme_minimal() +
  theme(legend.position = "bottom")
```

@@

Table \ref{tab:swn-results} replicates the first three models of SWN (the fourth model did not have any significant variables; see Table 1 in SNW). It is possible to verify that the results match with the minor (and irrelevant) exception of the magnitude of the coefficient for travel by private transportation, which is due to a difference in the input (here the variable is one percent units, whereas in SWN it was ten percent units). The intercept in these models is a random variable, and the standard deviation is reported in the fourth row of Table \ref{tab:swn-results}. It is useful to map the random intercepts: as seen in Figure \ref{fig:random-terms-map}, other things being equal, counties in Texas tended to have somewhat lower values of $R_0$ (i.e., a negative random intercept), whereas counties in South Dakota tended to have higher values of $R_0$. The key of the analysis, that population density has a positive association with the basic reproductive number, is verified. But is it really?

@@

```{r join-random-intercepts, include=FALSE}
state_geo <- state_geo %>% 
  left_join(model3$coefficients$random$state %>%
              as.data.frame() %>%
              rownames_to_column(var = "state"),
            by = "state") %>%
  rename(random_intercepts = `(Intercept)`)
```

@@

```{r random-terms-map, echo=FALSE, out.width= "1\\linewidth", fig.cap="\\label{fig:random-terms-map}Random intercepts of Model 3 (Alaska, Hawaii, Puerto Rico, and territories not shown)."}
# Choropleth map of basic reproductive number
ggplot() + 
  geom_sf(data = state_geo %>%
            filter(!STATE %in% c("02", "15", "72")), # Exclude Alaska, Hawaii, and Puerto Rico
          aes(fill = random_intercepts), # Add the simple features object to the plot
          color = "lightgray") +
  # Use viridis color palette: viriis color palettes are designed to print well in monochrome, are perceptually uniform, and effective for colorblindness
  scale_fill_viridis(name = "Random Intercepts",
                     option = "turbo",
                       direction = -1) +
  # Set theme of plot
  theme_minimal() +
  theme(legend.position = "bottom")
```

# Expanding on SWN

Thanks to the availability of code and data, it was possible to verify the results reported by SWN. As noted earlier, though, an experienced spatial modeller would have asked themselves about the implications of the spatial sampling procedure used by SWN. The decision to use a sample of counties with reliable basic reproductive numbers, although seemingly sensible, results in a non-random sampling scheme. Turning our attention back to Figure \ref{fig:R0-map}, we see that many counties without reliable values of $R_0$ are in more rural, less dense parts of the United States. The fact that $R_0$ could not be accurately computed in those counties does not mean that there was no transmission of the virus: it simply means that we do not know with precision whether that was the case. The low number of cases may be related to low population and/or low population density. This is intriguing, to say the least: by excluding cases in a non-random way such as this, we are in effect _truncating_ the sample.

The issue with sample truncation is that parameter estimates can become unreliable.

@@

```{r include=FALSE}
#
tobit1 <- censReg(R ~ density_log, 
                        left = 0,
                    data = county_geo_clean)

#
tobit2 <- censReg(R ~ density_log + private, 
                    data = county_geo_clean)

#
tobit3 <- censReg(R ~ density_log + private + hincome, 
                    data = county_geo_clean)
```

```{r include=FALSE}
tobit1_me <- tobit1 %>%
  margEff() %>%
  summary() %>%
  as.data.frame()

colnames(tobit1_me) <- c("me", "se", "t", "p")

tobit1_me <- tobit1_me %>%  rownames_to_column(var = "Variable") %>% 
  transmute(Variable,
            me,
            ll = me - 1.96 * se, 
            ul = me + 1.96 * se)

tobit2_me <- tobit2 %>%
  margEff() %>%
  summary() %>%
  as.data.frame()

colnames(tobit2_me) <- c("me", "se", "t", "p")

tobit2_me <- tobit2_me %>%  rownames_to_column(var = "Variable") %>% 
  transmute(Variable,
            me,
            ll = me - 1.96 * se, 
            ul = me + 1.96 * se)

tobit3_me <- tobit3 %>%
  margEff() %>%
  summary() %>%
  as.data.frame()

colnames(tobit3_me) <- c("me", "se", "t", "p")

tobit3_me <- tobit3_me %>%  rownames_to_column(var = "Variable") %>% 
  transmute(Variable,
            me,
            ll = me - 1.96 * se, 
            ul = me + 1.96 * se)

results_me_tobit <- tobit1_me %>%
  full_join(tobit2_me,
            by = "Variable") %>%
  full_join(tobit3_me,
            by = "Variable")

colnames(results_me_tobit) <- c("Variable",
                                "b1",
                                "l1",
                                "u1",
                                "b2",
                                "l2",
                                "u2",
                                "b3",
                                "l3",
                                "u3")

results_me_tobit <- results_me_tobit %>%
  transmute(Variable,
            b1 = ifelse(is.na(b1),
                        " ", 
                        round(b1, 3)),
            ci1 = ifelse(is.na(l1), 
                         " ",
                         paste0("[", round(l1, 3), ", ", round(u1, 3), "]")),
            b2 = ifelse(is.na(b2),
                        " ", 
                        round(b2, 3)),
            ci2 = ifelse(is.na(l2), 
                         " ",
                         paste0("[", round(l2, 3), ", ", round(u2, 3), "]")),
            b3 = ifelse(is.na(b3),
                        " ", 
                        round(b3, 3)),
            ci3 = ifelse(is.na(l3), 
                         " ",
                         paste0("[", round(l3, 3), ", ", round(u3, 3), "]")))
```

```{r echo=FALSE}
results_me_tobit %>%
  mutate(Variable = c("Log of population density",
                      "Percent of private transportation", 
                      "Median household income ($10,000)")) %>%
  kable("latex",
        digits = 3,
        booktabs = TRUE,
        col.names = c("Variable",
                      "m.e.",
                      "95% CI",
                      "m.e.",
                      "95% CI",
                      "m.e.",
                      "95% CI"),
        caption = "\\label{tab:tobit-results}Marginal effects of sample selection models") %>%
  kable_styling(latex_options = c("scale_down")) %>%
  add_header_above(c(" ", "Tobit Model 1" = 2, "Tobit Model 2" = 2, "Tobit Model 3" = 2))
```

@@

Fit spatially autoregressive tobit:
```{r include=FALSE}
# Fit SAR Tobit
fit_sartobit <- sartobit(R ~ density_log + private + hincome_log,
                         B,
                         ndraw = 1000,
                         burn.in = 200, 
                         showProgress = TRUE,
                         data = county_geo_clean,
                         computeMarginalEffects = TRUE)
summary(fit_sartobit)
impacts(fit_sartobit)
```
